---
title: "PS270 Final Project: Voter Turnout Differences between Urban and Rural States"
author: Antonietta Zamarra
description: "How does voter turnout differ between less densely populated states and more densely populated states?"
output:
  distill::distill_article:
    self_contained: false
---
 


## Introduction
My research question will be: "Do rural states have lower voter turnout overall than urban states?" My hypothesis related to this research question is that rural states, due to less dense populations, have less voter turnout overall than urban states, where populations are more dense. This research question is important because knowing the states in which less people vote will help improve future election engagement. A disparity in voter turnout will also lead to a lack of representation for specific demographics or interests, so if rural voter turnouts are less on average, rural states will be misrepresented in election outcomes. This makes it necessary to understand and address turnout disparities and their causes to ensure accurate representation in elections for all populations.

This research question has multiple possible confounders. One of these could be socioeconomic conditions, which may differ in urban and rural states and affect voter turnout. Education levels are one of these conditions, as urban states may have more residents with bachelor's degrees than rural states, which could lead to how important one perceives elections to be. Another confounder may be politician campaign activity, as politicians may concentrate more campaign efforts in urban states due to larger populations and more electoral votes. People also feel more encouraged to vote when they are mobilized through campaign efforts. Lower levels of campaign activity in more rural states results in reduced voter mobilization, leading to less political awareness and less perceived importance of an election. Individuals will therefore be less likely to participate in an election.



## Data Sources
```{r Election Turnout Data, include= FALSE}
library(tidyverse)

turnout <- read_csv("https://election.lab.ufl.edu/data-downloads/turnoutdata/Turnout_1980_2022_v1.2.csv")

head(turnout)
```


```{r State Population Density, include= FALSE}
pop_density <- read_csv("state_population_density_2020.csv")
head(pop_density)
```
My dataset for education levels is from the Federal Reserve Bank of St. Louis (FRED, 2021), including the percentage of bachelor degrees by state in 2021. 

I utilized a data set from the Federal Reserve Bank of St. Louis that provided the percentage of people with bachelor's degrees in 2021, to discuss the confounding variable of education levels by state (FRED, 2021). I also used a data set from the MIT Election Data and Science Lab with election turnout data from all 50 states in the last 4 elections (MIT Election Data and Science Lab, 2020). My data set with the population density of all 50 states comes from the United States Census Bureau (United States Census Bureau, 2020).

The independent variable is population density of a state, as states with the lowest population density will be designated as rural states, and states with the highest density will be labeled as urban states. The dependent variable is voter turnout, and this is measured as the percentage of registered voters who cast a ballot in each state in a given election year. I aim to prove that rural states (less densely populated) have lower voter turnouts in elections than urban states (more densely populated). This research design is cross-sectional and observational, since it involves analysis of existing data about polling places and election turnout at a specific time. Observations that support my hypothesis would be if urban and more densely populated states have higher voter turnout on average than less densely populated states. My hypothesis would be disproved if there is no positive relationship between population density and voter turnout.


### Variables

```{r State Density Chart, echo= FALSE}
pop_density_top10 <- pop_density |>
  slice_max(pop_density_2020, n = 10)

pop_density_top10 |>
  ggplot(aes(x = reorder(state, pop_density_2020), y = pop_density_2020)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "10 States with Highest Population Density (2020)",
       x = "State",
       y = "Population Density") +
  theme_minimal() 
```
 
This visualization shows the ten most densely populated states in the United States. The least densely populated states are considered rural in this study, while the states with the highest density are considered urban.

```{r Turnout (Dependent Variable) Chart, fig.height = 8, echo= FALSE}

turnout_2020 <- turnout |>
  filter(YEAR == 2020) |>
  filter(STATE != "United States") |>
  mutate(VEP_TURNOUT_RATE = as.numeric(gsub("%", "", VEP_TURNOUT_RATE)))

turnout_2020 |>
  filter(!is.na(VEP_TURNOUT_RATE)) |>
  ggplot(aes(x = reorder(STATE_ABV, VEP_TURNOUT_RATE), y = VEP_TURNOUT_RATE)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Voter Turnout by State (2020)",
       x = "State (Abbreviation)",
       y = "Voter Turnout (%)") +
  theme_minimal()
```
This bar chart shows voter turnout, in percent, for the 2020 election in all 50 states. There is clear variation across states in turnout percentages, as some states had a turnout rate above 75%, while some rates fall under 60%.
 
```{r State-by-State Election Turnout for Last Four Election Cycles, include=FALSE}
library(dplyr)

turnout_recent <- turnout |>
  filter(YEAR %in% c(2008, 2012, 2016, 2020))
```

```{r Sort Rural/Urban States, include=FALSE} 
turnout_recent <- turnout_recent |>
  mutate(state_type = ifelse(STATE_ABV %in% c("WY","VT","ND","MT","SD","AK","NH"), 
                            "Rural", "Urban"))
```

## Results
```{r Variable Scatterplot, echo= FALSE}

density_turnout <- pop_density |>
  inner_join(turnout_2020, by = c("state" = "STATE"))

ggplot(density_turnout, aes(x = pop_density_2020, y = VEP_TURNOUT_RATE)) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Population Density vs. Voter Turnout (2020)",
       x = "Population Density",
       y = "Voter Turnout (%)") +
  theme_minimal()
```
This scatterplot displays the slight positive trend between population density and voter turnout across states in 2020. The trend suggests more densely populated states do tend to have a slightly higher turnout than less densely populated states, though there are multiple outliers that are further away from the line. The wide spread of data points and the number of outliers suggest that population density alone cannot account for voter turnout, and there are likely other variables that influence the outcome.

```{r Results Line Plot, echo= FALSE} 
library(modelsummary)
library(broom)


turnout_recent <- turnout_recent |>
  mutate(VAP_TURNOUT_RATE = as.numeric(gsub("%", "", VAP_TURNOUT_RATE)))

turnout_summary <- turnout_recent |>
  group_by(YEAR, state_type) |>
  summarize(AvgTurnout = mean(as.numeric(VAP_TURNOUT_RATE), na.rm = TRUE), .groups = "drop")

ggplot(turnout_summary, aes(x = as.numeric(YEAR), y = AvgTurnout, color = state_type, group = state_type)) +
  geom_line(size = 1, alpha = 0.8) + 
  geom_point(size = 1, alpha = 0.8) +
  scale_color_manual(values = c("Rural" = "#F8766D", "Urban" = "#00BFC4")) +
  theme_minimal() +
  labs(
    y = "Average Turnout", 
    x = "Year", 
    color = "State Type")

```
```{r Regression, echo=FALSE}
model <- lm(VEP_TURNOUT_RATE ~ pop_density_2020, data = density_turnout)
tidy(model) |>
  select(term, estimate, std.error, p.value) |>
  mutate(term = case_when(
  term == "pop_density_2020" ~ "Population Density (per sq. mile)",
  term == "(Intercept)" ~ "Intercept"
)) |>
  knitr::kable(digits = 3, col.names = c("Term", "Estimate", "Std. Error", "p-value"))
```
The line plot shows the average voter turnouts in a given election year between rural and urban states. The results shown do not support my initial hypothesis because they show a higher turnout in rural states than urban states in the last 6 election cycles, meaning that a state being rural does not correspond with having lower voting turnout rates. The regression shows an intercept of 66.30, which represents a predicted baseline voter turnout of 66% when population density is zero. The coefficient of  0.0027 means that turnout increases by 0.003 percentage points for every one-unit increase in population density. Due to the p-value of 0.385, the relationship between a state's population density and voter turnout is not statistically significant. Despite the coefficient being positive, 0.003 percentage points per unit is a very small increase, showing that the effect that population density has on voter turnout is extremely minimal.

The difference in educational levels in an urban state compared to a rural state is a confounder that could affect voter turnout. Those with bachelor's degrees may have a better idea of election importance than those without, and an urban state may have more people with higher education levels than a rural state. Running a regression including the percentage of bachelor's degrees in each state helps control for this confounder.
```{r Attainment of Bachelors Degrees by State 2021, include=FALSE}
bach_degrees <- read_csv("educational_attainment_2021.csv")
head(bach_degrees)
```

```{r Confounder Regression, include= FALSE}

bach_degrees <- bach_degrees |>
  rename(bach_degrees = `2021`) |>
  mutate(bach_degrees = as.numeric(bach_degrees)) 

turnout_edu <- density_turnout |>
  inner_join(bach_degrees, by = c("state" = "State")) 

model_2 <- lm(VEP_TURNOUT_RATE ~ pop_density_2020 + bach_degrees, data = turnout_edu)

```
 
```{r Tidy Regression, echo=FALSE}
tidy(model_2) |>
  select(term, estimate, std.error, p.value) |>
  mutate(term = case_when(
  term == "pop_density_2020" ~ "Population Density (per sq. mile)",
  term == "(Intercept)" ~ "Intercept",
  term == "bach_degrees" ~ "Percentage of Bachelor's Degrees"
)) |>
  knitr::kable(digits = 3, col.names = c("Term", "Estimate", "Std. Error", "p-value"))

```
This multiple regression shows an intercept of 40.7, meaning that predicted voter turnout is around 41% when a state's population density and percentage of bachelor degrees are zero. This is a lower intercept than the initial regression, showing that level of education did account for some of the turnout baseline. The coefficient of -0.005 for population density shows that the effect of population density becomes negative when taking percentage of bachelor's degrees into consideration. The p-value of 0.053 is narrowly above the threshold of 0.05, which is not statistically significant. The percentage of bachelor degrees coefficient of 0.800 shows that when the amount of bachelor's degrees increases by 1%, voter turnout increases by 0.8 percentage points, revealing education level to be a more accurate predictor of voter turnout than state population density.
 



## Conclusion
The scatterplot displayed a slight positive correlation between population density and voter turnout, suggesting that states with higher population density trend towards higher voter turnout. However, the regression analysis showed an extremely small coefficient and a p-value of 0.385, indicating that the relationship between variables was not statistically significant. This means that the null hypothesis cannot be rejected, and a relationship between a state's voter turnout and population density could be due to chance. This relationship should not be interpreted causally because this study is cross-sectional and observational, and the population density of a state is not randomly assigned, so there may be existing confounders that explain the relationship between density and voter turnout.

These results do not support my hypothesis, because they do not prove that more densely populated states have higher voter turnout than states with lower population density. Some confounders to my hypothesis were campaign activity differences throughout states, and socioeconomic factors such as differing education rates. One limitation was that 2 states were dropped from my regression analysis, most likely due to a difference in names between my density and turnout datasets that I joined for two states. In addition, I was limited by only studying data from the 2020 election. I would improve this research in the future by studying data from multiple election years, or controlling for a different confounder, such as differences in other socioeconomic conditions across states. When controlling for education levels, the relationship between population density and voter turnout was shown to be negative in the multiple regression output. This suggests that education levels may be a more precise predictor of voter turnout than solely relying on population density.

## References
United States Elections Project. (2022). Voter turnout data (1980–2022). University of Florida. https://election.lab.ufl.edu/data-downloads/turnoutdata/

U.S. Census Bureau. (2020). Population density by state. United States Census Bureau.

Federal Reserve Bank of St. Louis. (2021). Share of population with a bachelor's degree or higher by state. FRED Economic Data. https://fred.stlouisfed.org

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}
```

